\class : Problem Set \psetnum

### 2022-05-08 531_stars/pset3/solutions.tex
stars pset3 q4

\question{\textbf{17.1 - Helium Flash Duration}}
\answer{}

>>>>

\clearpage

\question{\textbf{17.1 - Helium Flash Duration}}
\answer{
    First we need to calculate the potential energy of a uniform sphere. We know that gravitational potential is given by
    \begin{equation}
        \Phi = - \frac{G M}{r}
    \end{equation}
    So we just need to integrate this over a series of teeny tiny shells over the core of the star. Note we can assume a uniform density so $\rho(r) = \rho$. Let's do it!
    \begin{align}
        U &= - \int_0^R \frac{G M(r)}{r} \rho(r) \dd[3]{r} \\
          &= - \int_0^R \frac{G (\rho(r) \frac{4}{3} \pi r^3)}{r} \rho(r) 4 \pi r^2 \dd{r} \\
          &= - \frac{16 G \pi^2 \rho^2}{3}\int_0^R r^4 \dd{r} \\
        U &= - \frac{16 G \pi^2 \rho^2 R^5}{15}
    \end{align}
    We can rearrange this to get rid of the radius and just write it as a function of density and mass.
    \begin{align}
        R &= \qty(\frac{3 M}{4 \pi \rho})^{1/3} \\
        U &= - \frac{16 G \pi^2 \rho^2}{15} \qty(\frac{3 M}{4 \pi \rho})^{5/3} \\
        U &= - \frac{G \qty(36 \pi M^5 \rho)^{1/3}}{5}
    \end{align}
    Nice, what a...clean expression?\footnote{To be fair, it looks much better if we did it in terms of $M$ and $R$: $U = -\frac{3}{5} G M^2 / R$}. Now let's use this to find the difference in potential energy for a $0.5 \unit{M_\odot}$ core expanding from a density of $\rho = 10^{6} \unit{g}{cm^{-3}}$ to $\rho = 10^{4} \unit{g}{cm^{-3}}$.
    \begin{equation}
        \Delta U = - \frac{G \qty(36 \pi (0.5 \unit{M_\odot})^5)^{1/3}}{5} \qty[(10^{4} \unit{g}{cm^{-3}})^{1/3} - (10^{6} \unit{g}{cm^{-3}})^{1/3}] = 5 \times 10^{42} \unit{J}
    \end{equation}
    Now we can calculate the duration of the Helium flash given the energy production and efficiency
    \begin{equation}
        \tau_{\rm flash} = \frac{\Delta U}{L \cdot \phi} = \frac{5 \times 10^{49} \unit{J}}{10^{10} \unit{L_\odot} \cdot 0.2} = 2.5 \times 10^{40} \unit{erg}{L_\odot^{-1}}
    \end{equation}
    Now because I'm an ``astronomer'' I can 100\% (totally) understand ergs and $L_\odot$ intuitively, but just in case some readers don't live and breathe cgs in the same way this translates to
    \begin{equation}
        \boxed{ \tau_{\rm flash} = 0.21 \unit{yr} }
    \end{equation}

}
